Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Mechanical Perturbations

### 2024-05-23 main.tex
Added two papers to intro and improved wording.

\section{Introduction}
%
Single-actor bicycle crashes are associated with a suprisingly large number of
reported injuries~\citep{Wegman2024}. At low speeds, from start up to typical
cruising speeds, bicycles themselves are not self-stable and can be challenging
for the rider to balance. Low speed crashes may be reduced if the bicycle was
self-stable at these speeds, possibly relieving the rider of a necessary
control activity. Bicycles can be mechanically modified to lower the speeds at
which they are self-stable~\citep{Astrom2005} but since at least the 1980s it
is known that applying a steering torque proportional to the roll rate can
stabilize a single track vehicle at very low speeds. If the robotic control of
steering that implements this can stabilize a bicycle, it may reduce the
control required by the rider to successfully manage balancing tasks. We have
developed such a balance assisting bicycle and hypothesize that it helps the
rider in situations they are likely to fall.

Automatic roll stabilization of single track vehicles began in earnest after
predictive motorcycle models were developed and refined in throughout the
1970s. \Citet{vanZytveld1975} seems to be the first \todo{Check if a Japanese
researcher was first} to attempt to robotically stabilize a small motorbike
with a controlled inverted pendulum that mimicked rider lean, but he was not
successful in demonstrating what his theoretic control model predicted. In his
model, he recognized that feedback of the vehicle roll angle and angular rate
was essential to stabilize the vehicle. It was not until the mid 1980s that
someone successfully demonstrated an automatically balanced
motorcycle~\citep{Ruijs1986a} and they did so with a steering motor. Ruijs and
Pacejka clearly showed that steer torque driven by roll angle feedback
stabilizes the capsize mode, roll angular rate feedback stabilizes the weave
mode, and steer angular rate feedback stabilizes the wobble mode. They also
propose how the gains need to change with respect to vehicle speed for
favorable control at all speeds. This roll motion feedback is the simplest
controller that can stabilize a single track vehicle above a minimum speed when
not concerned with wobble instabilities, which are not prominent in bicycles at
low speeds. Ruijs and Pacejka's work was not particularly concerned with low
speed stability and their vehicle was fully automatic, no human rider was
involved.  Many more automatically balanced single track vehicles have been
demonstrated over the last 40 years, but none have demonstrated that increasing
low speed stability can assist in balancing and possibly reduce single actor
falls. Most of these robotic bicycles and motorcycles designers did not intend
for a human rider to also control the stabilized vehicle. Nevertheless, an
automatically stabilized bicycle can be controlled by a human rider if the
motor controlled steer torque and the rider applied steer torque act on the
steer in parallel. The effect of this automatic control gives the ability to
change the dynamics, up to some limits, of the human controlled plant.

\todo[inline]{We should cite Leila's first paper and also the Pforzhiem work
(if there is something to cite).}

The linear Carvallo-Whipple bicycle model~\cite{Carvallo1899,Whipple1899} is
the simplest bicycle model that exhibits non-minimum phase behavior and
self-stablity. Whipple showed that roll motion feedback can stabilize the
system. The model is ideal for showing the effect of a roll motion feedback
driven steer motor on the dynamics. The linear model can be described by the
fourth order state space equations:

@@

The state matrix \(\mathbf{A}\) and input matrix \(\mathbf{B}\) are both
functions of the equilibrium speed \(v\) and \(k_{\dot{\phi}}\) can be selected
such that the eigenvalues of \(\left(\mathbf{A} - \mathbf{B}_\delta\left[0 \quad k_{\dot{\phi}}
\quad 0 \quad 0\right] \right)\) have negative real parts for
\(v_{weave} < v < v_{capsize}\). With gain scheduling with respect to \(v\),
the speed range where the bicycle is stable can be maximized within
any physical acutator magnitude and bandwidth limits. \citet{Schwab2008}
elaborates on some of the possibilities in scheduling the gains for such a
controller.

@@

\section{Methods}
%
\subsection{Bicycle}
%
We modified a standard electric bicycle (Gazelle XXX, Dieren, The Netherlands)
with a custom motor in the steering assembly capable of applying up to
7~\si{\newton\meter} of torque between the head tube and steer tube. A custom
motor controller converts the commanded torque to applied torque. We measure
the speed of the rear wheel with an encoder (Magura XXX) and measure the body
fixed roll rate of the bicycle with a MEMs rate gyroscope (Company, Location).
The balance assist control algorithm is implemented on a microprocessor
(Teensy, Company, Location) and data from all sensors is logged with a CAN bus
(Company, Location) at approximately X Hz.

@@

\subsection{Perturbations}
%
We apply external longitudinal forces to the ends of each handlebar using an
adapted Bump'Em system~\cite{Tan2020} which is designed with four motors
working in tandem. The four motors are programmed to apply a light force at all
times to keep the ropes taught and to track a commanded force profile using a
PID controller running on a microprocessor (Arduino XXX, Italy). We measure the
force applied by each motor at the handlebar via four inline load cells, each
with a max load of XX~\si{\newton}. The commanded force profiles are designed
to apply an external pulsive torque to the front assembly (handlebars, forks,
wheel) at magnitudes varying from X to X. The four motors are arranged at the
four corners of a 1~\si{\meter} wide treadmill (make, Company, City) that can
reach speeds of 18~\si{\kilo\meter\per\hour}. The general design is described
in detail in the Van De Velde's MSc thesis~\cite{vandeVelde2022}. Our
modifications relative to that described in the thesis include simplifying the
controller with a inexpensive microcontroller and the use of a simpler safety
harness.

@@

Based on the findings in \cite{Haitjema2023}, we calculate several variables.
We use the angular impulse \(L\) due to the perturbation forces over a
0.3~\si{\second} duration to characterize the magnitude of delivered
perturbation. The duration is selected based on the duration of the commanded
force.

>>>>

\section{Introduction}
%
Single-actor bicycle crashes are associated with a suprisingly large number of
reported injuries~\citep{Wegman2024}. At low speeds, from start up to typical
cruising speeds, bicycles themselves are not self-stable and can be challenging
for the rider to balance. Low speed crashes may be reduced if the bicycle was
self-stable at these speeds, possibly relieving the rider of a necessary
control activity. Bicycles can be mechanically modified to lower the speeds at
which they are self-stable~\citep{Astrom2005} but since the 1980s it has been
known that applying a motor acutated steering torque proportional to the roll
rate can stabilize a single track vehicle at very low speeds. If this automatic
control of steering can stabilize a bicycle, it may reduce the control required
by the rider to successfully manage balancing tasks. We have developed such a
balance assisting bicycle and hypothesize that it helps the rider in situations
they are likely to fall.

Automatic roll stabilization of single track vehicles began in earnest after
predictive motorcycle models were developed and refined in throughout the
1970s. \Citet{vanZytveld1975} seems to be the first \todo{Check if a Japanese
researcher was first} to attempt to robotically stabilize a small motorbike
with a controlled inverted pendulum that mimicked rider lean, but he was not
successful in demonstrating what his theoretic control model correctly predicted. In his
model, he recognized that feedback of the vehicle roll angle and angular rate
was essential to stabilize the vehicle. It was not until the mid 1980s that
Ruijs and Pacjecka successfully demonstrated an automatically balanced
motorcycle~\citep{Ruijs1986} and they did so with a steering motor. Ruijs and
Pacejka clearly showed that steer torque driven by roll angle feedback
stabilizes the capsize mode, roll angular rate feedback stabilizes the weave
mode, and steer angular rate feedback stabilizes the wobble mode. They also
propose how the gains must change with respect to vehicle speed for
favorable control at all speeds. This roll motion feedback is the simplest
controller that can stabilize a single track vehicle above a minimum speed when
not concerned with wobble instabilities, which are not prominent in bicycles at
low speeds. Ruijs and Pacejka's work was not particularly concerned with low
speed stability and their vehicle was fully automatic, no human rider was
involved. Many more automatically balanced single track vehicles have been
demonstrated over the last 40 years, but none have demonstrated that increasing
low speed stability can assist in balancing and possibly reduce single actor
falls. Most of these robotic bicycles and motorcycles designers did not intend
for a human rider to also control the stabilized vehicle. Nevertheless, an
automatically stabilized bicycle can be controlled by a human rider if the
motor controlled steer torque and the rider applied steer torque act on the
steer in parallel. The effect of this automatic control gives the ability to
change the dynamics, up to some limits, of the human controlled plant.In our
prior study, we demonstrate reduced motion during
distractions~\cite{Alizadehsaravi2023} due to the balance assist but
researchers with a similar vehicle recently showed rider disatisfction with the
stabilization~\cite{Hanakam2023}.

The linear Carvallo-Whipple bicycle model~\citep{Carvallo1899,Whipple1899} is
the simplest bicycle model that exhibits non-minimum phase behavior and
self-stablity. Whipple even showed that roll motion feedback can stabilize the
system, remarkably in the early years of developments automatic control. The
bicycle model is ideal for showing the effect of a roll motion feedback driven
steer motor on the dynamics. The linear model can be described by the fourth
order state space equations:

@@

The state matrix \(\mathbf{A}\) and input matrix \(\mathbf{B}\) are both
functions of the equilibrium speed \(v\) and \(k_{\dot{\phi}}\) can be selected
such that the eigenvalues of \(\left(\mathbf{A} - \mathbf{B}_\delta\left[0 \quad k_{\dot{\phi}}
\quad 0 \quad 0\right] \right)\) have negative real parts for
\(v_{weave} < v < v_{capsize}\). With gain scheduling with respect to \(v\),
the speed range where the bicycle is stable can be maximized within
any physical acutator magnitude and bandwidth limits. \citet{Schwab2008}
elaborates on some of the possibilities in scheduling the gains for such a
controller on this model.

@@

\section{Methods}
%
\subsection{Bicycle}
%
We modified a standard electric bicycle (Gazelle Grenoble C7, Dieren, The
Netherlands) with a custom motor in the steering assembly capable of applying
up to 7~\si{\newton\meter} of torque between the head tube and steer tube. A
custom motor controller converts the commanded torque to applied torque. We
measure the speed of the rear wheel with an encoder (Magura XXX) and measure
the body fixed roll rate of the bicycle with a MEMs rate gyroscope (Company,
Location).  The balance assist control algorithm is implemented on a
microprocessor (Teensy, Company, Location) and data from all sensors is logged
with a CAN bus (Company, Location) at approximately X Hz.

@@

\todo[inline]{Decide on what physical parameters to use.}

\subsection{Perturbations}
%
We apply external longitudinal forces to the ends of each handlebar using an
adapted Bump'Em system~\cite{Tan2020} which is designed with four motors
working in tandem. The four motors are programmed to apply a light force at all
times to keep the ropes taught and to track a commanded force profile using a
PID controller running on a microprocessor (Arduino XXX, Italy). We measure the
force applied by each motor at the handlebar via four inline load cells, each
with a max load of XX~\si{\newton}. The commanded force profiles are designed
to apply an external pulsive torque to the front assembly (handlebars, forks,
wheel) at magnitudes varying from X to X. The four motors are arranged at the
four corners of a 1~\si{\meter} wide treadmill (make, Company, City) that can
reach speeds of 18~\si{\kilo\meter\per\hour}. The general design is described
in detail in the Van De Velde's MSc thesis~\cite{vandeVelde2022a}. Our
modifications relative to that described in the thesis include simplifying the
controller with a inexpensive microcontroller and the use of a simpler safety
harness.

@@

Based on the findings in \cite{Haitjema2023a}, we calculate several variables.
We use the angular impulse \(L\) due to the perturbation forces over a
0.3~\si{\second} duration to characterize the magnitude of delivered
perturbation. The duration is selected based on the duration of the commanded
force.
